\TheTitle

### 2019-07-15 Paper1/introduction.tex
continued paper

Due to the randomness of the solution, one is interested in determining the expectation value or the variance, i.e.
\begin{align*}
\text{E}[\bm{u}] = \langle \bm{u} \rangle,\qquad \text{Var}[\bm{u}] = \langle \left( \bm{u}-\text{E}[\bm{u}]\right)^2\rangle,
\end{align*}
where we use the bracket operator $\langle \cdot \rangle := \int_{\Theta} \cdot f_{\Xi}(\bm\xi)d\xi_1 \cdots d\xi_s$. To approximate quantities of interest (such as expectation value, variance or higher order moments), the solution is spanned with a set of polynomial basis functions $\varphi_{i}:\Theta\to\mathbb{R}$ such that for the multi-index $i = (i_1,\cdots,i_s)$ we have $|i| \leq M$. Note that this yields
\begin{align*}
N:=\begin{pmatrix}
M+s \\ s
\end{pmatrix}
\end{align*}
basis functions when defining $|i|:=\sum_{n = 1}^s |i_n|$. Commonly, these functions are chosen to be orthonormal polynomials \cite{wiener1938homogeneous} with respect to the probability function, i.e. $\langle \varphi_i \varphi_j \rangle =\prod_{n=1}^s\delta_{i_nj_n}$. The generalized polynomial chaos (gPC) expansion \cite{xiu2002wiener} approximates by the solution by
\begin{align}\label{eq:SGClosure}
\mathcal{U}(\bm{\hat u};\bm\xi):= \sum_{|i|\leq M} \bm{\hat{u}}_i\varphi_i(\bm{\xi}) = \hat{\bm u}^T\bm{\varphi}(\bm\xi),
\end{align}
where the deterministic expansion coefficients $\bm{\hat{u}}_i\in\mathbb{R}^p$ are called moments. To allow a more compact notation, we collect the $N$ moments for which $\vert i \vert \leq M$ holds in the moment matrix $\hat{\bm u}:=(\bm{\hat{u}}_i)_{|i\leq M|}\in\mathbb{R}^{N\times p}$ and the corresponding basis functions in $\bm{\varphi}:=(\varphi_i)_{|i|\leq M}\in\mathbb{R}^{N}$. In the following, the dependency of $\mathcal{U}$ on $\bm \xi$ will occasionally be omitted for sake of readability. The solution ansatz \eqref{eq:SGClosure} is $L^2$ optimal if the moments are chosen to be the Fourier coefficients $\bm{\hat u}_i := \langle \bm{u}\varphi_i \rangle\in\mathbb{R}^p$. The solution ansatz \eqref{eq:SGClosure} can then be used to compute the quantities of interest. Indeed, we have that
\begin{align*}
\text{E}[\mathcal{U}(\bm{\hat u})] = \bm{\hat u}_0,\quad \text{Var}[\mathcal{U}(\bm{\hat u})] = \text{E}[\mathcal{U}(\bm{\hat u})^2] - \text{E}[\mathcal{U}(\bm{\hat u})]^2 = \left(\sum_{i = 1}^N \hat{u}_{\ell i}^2\right)_{\ell = 1,\cdots,p}.
\end{align*}

Numerical methods for approximating the moment $\bm{\hat u}_i$ can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule: For a given set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Commonly, SC uses sparse grids as quadrature rule, since they posses a reduced number of collocation points for multi-dimensional problems. Compared to tensorized quadrature sets, which require $O(M^s)$ quadrature points, sparse grids use $O(M(\log_2(M)^{s-1}))$ quadrature points to integrate polynomials of total degree $M$ exactly. 
Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations can be carried out in parallel on different cores. A downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Therefore, collocation methods typically require more runs of the deterministic solver than intrusive methods \cite{xiu2009fast,alekseev2011estimation} to reach a given accuracy. Despite their easy implementation, collocation methods face challenges when examining unsteady problems, since in this case, the solution needs to be written out, i.e. stored in an external file, at all quadrature points in every time step to compute the time evolution of the moments. This contradicts modern HPC paradigms, which aim at reducing the amount of data produced by numerical methods.

@@

\bm{\hat u_i}(t=0,\bm{x}&) = \langle\bm{u}_{\text{IC}}(\bm{x},\cdot)\varphi_i\rangle

@@

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through the spatial numerical method \cite{offner2017stability} or filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{durrwachter2018hyperbolicity}. These structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. The main task here is to find an adequate refinement indicator. \\

In this paper, we present a semi-intrusive code framework, which facilitates the task of implementing general intrusive methods. The framework only requires the numerical flux of the deterministic problem (as well as an entropy if IPM is used). We thereby provide the ability to recycle existing implementations of deterministic solvers.
Furthermore, we investigate intrusive methods for steady problems and compare them against collocation methods. The steady setting provides different opportunities to take advantage of features of intrusive methods: 
\begin{itemize}
\item Accelerate the convergence to the IPM steady state solution by applying IPM as a post-processing step for collocation methods: We converge the moments of the solution to a steady state with an inaccurate, but cheap collocation method and then use the resulting collocation moments as starting values for an expensive but accurate intrusive method such as IPM, which we then again converge to steady state. 
\item Compute inexact dual variables \eqref{eq:dualProblem} for IPM: Since the moments during the iteration process are inaccurate, i.e. they are not the correct steady state solution, we propose to not fully converge the dual iteration, which computes \eqref{eq:dualProblem}. 
\end{itemize}
For both steady and unsteady problems, we use adaptivity: 
\begin{itemize}
\item The intrusive nature of SG and IPM can be used to locally increase the number of moments whenever the solution has a complex structure in the random variable (as well as decrease the number of moments if not). In the case of steady problems, one can perform a large number of iteration to the steady state solution on a low number of moments and increase the maximal refinement level when the distance to the steady state has reached a specified barrier. Similar to the idea of using IPM as a post processing step, a large number of iterations will be performed by a cheap method, i.e. we can reduce numerical costs. 
\end{itemize}
The effectiveness of these acceleration ideas are tested by comparing results with stochastic-Collocation for the uncertain NACA test case as well as a bent shock tube problem.

The paper is structured as follows: After the introduction, we discuss the numerical discretization as well as the implementation and structure of the semi-intrusive framework in section \ref{sec:framework}. Section \ref{sec:collIPM} discusses the IPM acceleration with a non-intrusive method and in section \ref{sec:OneShotIPM}, we discuss the idea of not converging the dual iteration. Section~\ref{sec:adaptivity} extends the presented numerical framework to an algorithm making use of adaptivity. A comparison of results computed with the presented methods is then given in \ref{sec:results}, followed by a summary and outlook in section \ref{sec:summary_outlook}.

>>>>

Due to the randomness of the solution, one is interested in determining the expectation value or the variance, i.e.
\begin{align*}
\text{E}[\bm{u}] = \langle \bm{u} \rangle,\qquad \text{Var}[\bm{u}] = \langle \left( \bm{u}-\text{E}[\bm{u}]\right)^2\rangle,
\end{align*}
where we use the bracket operator $\langle \cdot \rangle := \int_{\Theta} \cdot f_{\Xi}(\bm\xi)d\xi_1 \cdots d\xi_s$. To approximate quantities of interest (such as expectation value, variance or higher order moments), the solution is spanned with a set of polynomial basis functions $\varphi_{i}:\Theta\to\mathbb{R}$ such that for the multi-index $i = (i_1,\cdots,i_s)$ we have $|i| \leq M$. Note that this yields
\begin{align}\label{eq:numberBasisFcts}
N:=\begin{pmatrix}
M+s \\ s
\end{pmatrix}
\end{align}
basis functions when defining $|i|:=\sum_{n = 1}^s |i_n|$. Commonly, these functions are chosen to be orthonormal polynomials \cite{wiener1938homogeneous} with respect to the probability function, i.e. $\langle \varphi_i \varphi_j \rangle =\prod_{n=1}^s\delta_{i_nj_n}$. The generalized polynomial chaos (gPC) expansion \cite{xiu2002wiener} approximates the solution by
\begin{align}\label{eq:SGClosure}
\mathcal{U}(\bm{\hat u};\bm\xi):= \sum_{|i|\leq M} \bm{\hat{u}}_i\varphi_i(\bm{\xi}) = \hat{\bm u}^T\bm{\varphi}(\bm\xi),
\end{align}
where the deterministic expansion coefficients $\bm{\hat{u}}_i\in\mathbb{R}^p$ are called moments. To allow a more compact notation, we collect the $N$ moments for which $\vert i \vert \leq M$ holds in the moment matrix $\hat{\bm u}:=(\bm{\hat{u}}_i)_{|i\leq M|}\in\mathbb{R}^{N\times p}$ and the corresponding basis functions in $\bm{\varphi}:=(\varphi_i)_{|i|\leq M}\in\mathbb{R}^{N}$. In the following, the dependency of $\mathcal{U}$ on $\bm \xi$ will occasionally be omitted for sake of readability. The solution ansatz \eqref{eq:SGClosure} is $L^2$ optimal if the moments are chosen to be the Fourier coefficients $\bm{\hat u}_i := \langle \bm{u}\varphi_i \rangle\in\mathbb{R}^p$. The solution ansatz \eqref{eq:SGClosure} can then be used to compute the quantities of interest. Indeed, we have that
\begin{align*}
\text{E}[\mathcal{U}(\bm{\hat u})] = \bm{\hat u}_0,\quad \text{Var}[\mathcal{U}(\bm{\hat u})] = \text{E}[\mathcal{U}(\bm{\hat u})^2] - \text{E}[\mathcal{U}(\bm{\hat u})]^2 = \left(\sum_{i = 1}^N \hat{u}_{\ell i}^2\right)_{\ell = 1,\cdots,p}.
\end{align*}

Numerical methods for approximating the moment $\bm{\hat u}_i$ can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule: For a given set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Commonly, SC uses sparse grids as quadrature rule, since they posses a reduced number of collocation points for multi-dimensional problems. Compared to tensorized quadrature sets, which require $O(M^s)$ quadrature points, sparse grids use $O(M(\log_2(M)^{s-1}))$ quadrature points to integrate polynomials of total degree $M$ exactly. 
Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations can be carried out in parallel on different cores. A downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Therefore, collocation methods typically require more runs of the deterministic solver than intrusive methods \cite{xiu2009fast,alekseev2011estimation} to reach a given accuracy. Despite their easy implementation, collocation methods face challenges when examining unsteady problems. In this case, the solution needs to be written out, i.e. stored in an external file, at all quadrature points in every time step to compute the time evolution of the moments. This contradicts modern HPC paradigms, which aim at reducing the amount of data produced by numerical methods.

@@

\bm{\hat u_i}(t=0,\bm{x}&) = \langle\bm{u}_{\text{IC}}(\bm{x},\cdot)\varphi_i\rangle,

@@

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through the spatial numerical method \cite{offner2017stability} or filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{durrwachter2018hyperbolicity}. These structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. In addition to that, the number of unknowns when using intrusive instead of non-intrusive methods is significantly decreased in high dimensional problems: The number of moments $N$ given by \eqref{eq:numberBasisFcts} is asymptotically smaller than the number of quadrature points, which increases with $O(M(\log_2(M)^{s-1}))$. Therefore, one aim should be to accelerate intrusive methods, since they can potentially outperform non-intrusive methods in complex, high-dimensional settings. \\

In this paper, we investigate intrusive methods for steady problems and compare them against collocation methods. For both steady and unsteady problems, we use adaptivity: 
\begin{itemize}
\item The intrusive nature of SG and IPM can be used to locally increase the number of moments whenever the solution has a complex structure in the random variable (as well as decrease the number of moments if not).
\end{itemize}
The steady setting provides different opportunities to take advantage of features of intrusive methods: 
\begin{itemize}
\item When using adaptivity, one can perform a large number of iteration to the steady state solution on a low number of moments and increase the maximal refinement level when the distance to the steady state has reached a specified barrier. Consequently, a large number of iterations will be performed by a cheap method, i.e. we can reduce numerical costs. 
\item Accelerate the convergence to the IPM steady state solution by applying IPM as a post-processing step for collocation methods: We converge the moments of the solution to a steady state with an inaccurate, but cheap collocation method and then use the resulting collocation moments as starting values for an expensive but accurate intrusive method such as IPM, which we then again converge to steady state. 
\item Compute inexact dual variables \eqref{eq:dualProblem} for IPM: Since the moments during the iteration process are inaccurate, i.e. they are not the correct steady state solution, we propose to not fully converge the dual iteration, which computes \eqref{eq:dualProblem}. 
\end{itemize}

The effectiveness of these acceleration ideas are tested by comparing results with stochastic-Collocation for the uncertain NACA test case as well as a bent shock tube problem. Furthermore, we present a semi-intrusive method, which facilitates the task of implementing general intrusive methods. The method only requires the numerical flux of the deterministic problem (as well as an entropy if IPM is used). We thereby provide the ability to recycle existing implementations of deterministic solvers.

The paper is structured as follows: After the introduction, we discuss the numerical discretization as well as the implementation and structure of the semi-intrusive method in section \ref{sec:framework}. Section \ref{sec:collIPM} discusses the IPM acceleration with a non-intrusive method and in section \ref{sec:OneShotIPM}, we discuss the idea of not converging the dual iteration. Section~\ref{sec:adaptivity} extends the presented numerical framework to an algorithm making use of adaptivity. A comparison of results computed with the presented methods is then given in \ref{sec:results}, followed by a summary and outlook in section \ref{sec:summary_outlook}.

### 2019-07-15 Paper1/main.tex
continued paper

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. A main drawback of intrusive methods is the need to implement new code, whereas collocation methods recycle a given deterministic solver. Moreover, non-intrusive methods are embarrassingly parallel and therefore easy to parallelize.
In this work, we present a code framework, which facilitates the implementation of intrusive methods by recycling parts of deterministic codes. Additionally, we introduce methods to decrease numerical costs of intrusive methods for steady problems. We demonstrate the effectiveness of the proposed strategies by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment.
\end{abstract}

>>>>

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. A main drawback of intrusive methods is the need to implement new code, whereas collocation methods recycle a given deterministic solver. Unlike intrusive methods, non-intrusive methods yield a decoupled system of equations, meaning that they are embarrassingly parallel. Moreover, the coupled system of the intrusive stochastic-Galerkin (SG) method is not necessarily hyperbolic. The Intrusive Polynomial Moment (IPM) method fixes this issue at the cost of having to solve an optimization problem in every spatial cell and every time step.
In this work, we discuss advantages of intrusive methods and propose acceleration techniques for SG and IPM. When solving steady problems, the numerical costs arising from repeatedly solving the IPM optimization problem can be reduced by borrowing ideas from shape optimization. By integrating the iteration to solve the optimization problem into the moment update, we can guarantee local convergence while reducing numerical costs. Furthermore, we propose to use non-intrusive methods as a preconditioner for intrusive methods. Consequently, a large amount of iterations to the steady solution are performed by a cheap method, while maintaining the increased accuracy of an intrusive method. Additionally, we propose an adaptive implementation of the IPM method, which further reduces numerical costs. We demonstrate the effectiveness of the proposed strategies by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment.
\end{abstract}
